### 2015-04-22 doc/thesis_puc/typedlua.tex
Adding all chapters to PUC-Rio template

Typed Lua is an optional type system for Lua, and its main goal is to
provide static type checking for Lua.
To do that, Typed Lua extends the syntax of Lua 5.3 to introduce
optional type annotations, and performs local type inference \citep{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

@@

We use the consistent-subtyping relation of gradual typing
\citep{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
We did not insert run-time checks at this moment because they can decrease
run-time performance \citep{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done before
inserting them in the type system.
However, this evaluation is out of scope of this work.

Unlike Dart \citep{dart} and TypeScript \citep{typescript},
we are designing Typed Lua aiming soundness to make it possible to
switch Typed Lua from optional typing to gradual typing in the future.
A sound type system is a prerequisite to insert run-time checks after
static type checking, because a sound type system ensures that
statically typed code will not throw type errors during run-time.

@@

Typed Lua uses consistent-subtyping to check the interaction among the
dynamic type \texttt{any} and other types.
The dynamic type \texttt{any} is a subtype of \texttt{value}, but it is
neither a supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \citep{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and we can pass any value where a
value of the dynamic type is expected, but the compiler tracks these
operations, and the programmer can choose to be warned about them.
We will discuss the formalization of the consistent-subtyping
relation in Section \ref{sec:subtyping}.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the predicate \texttt{type(v) == "string"} that
constrains the type of \texttt{v} from \texttt{string|t} to
\texttt{string} when the predicate is true and \texttt{t} otherwise.
This is a simplified form of \emph{flow typing} \citep{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

\newpage
\section{Description files}
\label{sec:tld}

@@

The following example shows the description file for \texttt{lmd5} \citep{lmd5},
a MD5 digest library for Lua that is written in C:
\begin{verbatim}

>>>>

Typed Lua is an optional type system for Lua, and its main goal is to
provide static type checking for Lua.
To do that, Typed Lua extends the syntax of Lua 5.3 to introduce
optional type annotations, and performs local type inference \cite{pierce2000lti}
to detect more precise types for unannotated expressions.
Even though the compiler warns the programmer about type errors,
it always removes the type annotations to generate Lua code that
runs in unmodified Lua implementations.

@@

We use the consistent-subtyping relation of gradual typing
\cite{siek2007objects,siek2013mutable} to formalize Typed Lua,
though it does not insert run-time checks in the gradual typing style.
In gradual typing, run-time checks inspect the interaction between
dynamically typed and statically typed code to guarantee that dynamically
typed code does not violate statically typed code during run-time.
We did not insert run-time checks at this moment because they can decrease
run-time performance \cite{allende2013cis}.
We believe that a careful evaluation of run-time checks should be done before
inserting them in the type system.
However, this evaluation is out of scope of this work.

Unlike Dart \cite{dart} and TypeScript \cite{typescript},
we are designing Typed Lua aiming soundness to make it possible to
switch Typed Lua from optional typing to gradual typing in the future.
A sound type system is a prerequisite to insert run-time checks after
static type checking, because a sound type system ensures that
statically typed code will not throw type errors during run-time.

@@

Typed Lua uses consistent-subtyping to check the interaction among the
dynamic type \texttt{any} and other types.
The dynamic type \texttt{any} is a subtype of \texttt{value}, but it is
neither a supertype nor a subtype of any other type.
Our consistent-subtyping relationship follows the standards defined
by the gradual typing of objects \cite{siek2007objects,siek2013mutable}.
In practice, we can pass a value of the dynamic type anytime we want
a value of some other type, and we can pass any value where a
value of the dynamic type is expected, but the compiler tracks these
operations, and the programmer can choose to be warned about them.
We will discuss the formalization of the consistent-subtyping
relation in Section \ref{sec:subtyping}.

@@

Typed Lua has a small set of \texttt{type} predicates that allows
programmers to constrain the type of a local variable inside a condition.
This example uses the predicate \texttt{type(v) == "string"} that
constrains the type of \texttt{v} from \texttt{string|t} to
\texttt{string} when the predicate is true and \texttt{t} otherwise.
This is a simplified form of \emph{flow typing} \cite{guha2011tlc,tobin-hochstadt2010ltu}.
As with \texttt{or}, the variable must be local to the function
and it is not being assigned in another function.

@@

\section{Description files}
\label{sec:tld}

@@

The following example shows the description file for \texttt{lmd5} \cite{lmd5},
a MD5 digest library for Lua that is written in C:
\begin{verbatim}
